---
title: "S3 vectors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{S3 vectors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE, eval = FALSE}
# Find all methods that base provides
library(tidyverse)
vectors <- c("factor", "Date", "POSIXct", "difftime")

methods <- map_dfr(vectors, sloop::s3_methods_class)
methods %>% 
  filter(source == "base") %>% 
  View()
```


This vignette contains tips for creating your own S3 vector class. It focusses on the details of making a vector class play nice; you'll also need to provide methods that actually make the vector useful.

## Construction

All classes should have a low-level constructor called `new_class()` that checks types (but not values), then calls `structure()`. Every class should have one of `class()` and `as_class()`. Some classes will have both.

When called with no arguments the construct should return a zero-length vector.

## First methods
  
The first method you should implement is a `print()` method. Your first pass at a `print()` method should (ab)use existing `print()` methods. Your `print()` method should return `x` invisibly.

The second method you should implement is `[`. It should will look something like:

```{r}
`[.myclass` <- function(x, i) {
  new_myclass(NextMethod(), 
    attr1 = attr(x, "attr1"), 
    attr2 = attr(x, "attr2")
  )
}
```

If you are willing to take a dependency on sloop, you should implement a reconstruct method:

```{r}
reconstruct.myclass <- function(new, old) {
  new_myclass(new, 
    attr1 = attr(old, "attr1"), 
    attr2 = attr(old, "attr2")
  )
}
```

Then your `[` method can be simplified to:

```{r}
`[.myclass` <- function(x, i) {
  reconstruct(NextMethod(), x)
}
```

## Other methods

Implement `[<-` to ensure that the user can't corrupt the data type.

Consider implementing `names<-()` and `dims<-()` methods to expressly forbid use of names or dims (unless you actually want to support them).

Avoid redefining `length()` or `names()`. If you find yourself leading in this direction, it may be a sign that you need a vector built on top of a data frame instead.
  
## Data frames and tibbles
  
Next implement `as.data.frame.myclass` so you can put your vector inside a data frame. It should always have this form:

```{r}
as.data.frame.myclass <- function(x, row.names = NULL, optional = FALSE, ..., nm = paste(deparse(substitute(x), width.cutoff = 500L), collapse = " ")) {
  force(nm)
  
  cols <- list(x)
  if (!optional) {
    names(cols) <- nm
  }
  
  structure(
    cols,
    class = "data.frame",
    row.names = .set_row_names(length(x))
  )
}
```

If you need the column to print specially within the data frame provide a `format()` method. It should return a character vector the same length as `x`. Consider rewriting your print method to use `format()`.
  
To make work more nicely in tibbles, provide a method for 
`pillar::type_sum()`, and if you need custom printing, a method for
`pillar::pillar_shaft()`. See details in 
<https://tibble.tidyverse.org/articles/extending.html>.

## Utils

```{r}
cat_line <- function(...) cat(..., "\n", sep = "")
```

## Case study: percent (vctrs)

```{r}
library(vctrs)
new_percentv <- function(x) {
  stopifnot(is.double(x))
  
  new_vctr(x, class = "percentv")
}

percentv <- function(x = double()) {
  # Check invariants
  if (!is.numeric(x)) {
    stop("Percentages must be numeric", call. = FALSE)
  }
  if (any(x < 0 | x > 1, na.rm = TRUE)) {
    stop("Percentages must be between 0 and 1", call. = FALSE)
  }
  
  # Strip attributes and names
  x <- as.vector(x)
  names(x) <- NULL
  
  new_percentv(x)
}

format.percentv <- function(x, ...) {
  out <- format.default(signif(unclass(x) * 100, 3))
  out[!is.na(x)] <- paste0(out[!is.na(x)], "%")
  out
}

x <- percentv(c(seq(0, 1, length = 4), NA))
x

x[c(1, 1, 4)]
x[0]
x[[1]]

df <- data.frame(x = x)
df

type_sum.percentv <- function(x) {
  "pctv"
}

tibble::tibble(x)
```

## Case study: percent (base)

Foundations:

```{r}
new_percent <- function(x) {
  stopifnot(is.double(x))
  structure(x, class = "percent")
}

percent <- function(x = double()) {
  # Check invariants
  if (!is.numeric(x)) {
    stop("Percentages must be numeric", call. = FALSE)
  }
  if (any(x < 0 | x > 1, na.rm = TRUE)) {
    stop("Percentages must be between 0 and 1", call. = FALSE)
  }
  
  # Strip attributes and names
  x <- as.vector(x)
  names(x) <- NULL
  
  new_percent(x)
}

print.percent <- function(x, ...) {
  cat_line("<percent>")
  if (length(x) > 0) 
    print(paste0(signif(x * 100, 3), "%"), quote = FALSE)
  invisible(x)
}

x <- percent(c(seq(0, 1, length = 4), NA))
x

`[.percent` <- function(x, i) {
  new_percent(NextMethod())
}

x[c(1, 1, 4)]

# Now is a good time to check that zero-length vectors print well
x[0]

`[[.percent` <- function(x, i) {
  new_percent(NextMethod())
}

x[[1]]
```

Next make it work in data frames:

```{r}
as.data.frame.percent <- function(x, 
                                  row.names = NULL, 
                                  optional = FALSE, 
                                  ..., 
                                  nm = paste(deparse(substitute(x), width.cutoff = 500L), collapse = " ")
                                  ) {
  force(nm)
  
  cols <- list(x)
  if (!optional) {
    names(cols) <- nm
  }
  
  structure(
    cols,
    class = "data.frame",
    row.names = .set_row_names(length(x))
  )
}

df <- data.frame(x = x)
df

format.percent <- function(x, ...) {
  out <- paste0(format.default(signif(x * 100, 3)), "%")
  out[is.na(x)] <- NA
  out
}
df

# Now rewrite print
print.percent <- function(x, ...) {
  cat_line("<percent>")
  if (length(x) > 0) 
    print(format(x), quote = FALSE)
  invisible(x)
}
x
```

And in tibbles:

```{r}
type_sum.percent <- function(x) {
  "pct"
}

tibble::tibble(x)
```

Make it stricter:

```{r, error = TRUE}
`[<-.percent` <- function(x, i, value) {
  value <- percent(value)
  NextMethod()
}

`[[<-.percent` <- function(x, i, value) {
  value <- percent(value)
  NextMethod()
}

x[1] <- "a"
x[[1]] <- 2

x[1:2] <- 0.5
x

`names<-.percent` <- function(x, value) {
  stop("Percent vector must be nameless", call. = FALSE)
}

`dim<-.percent` <- function(x, value) {
  stop("Percent vector must be 1d", call. = FALSE)
}
```

