---
title: "vctrs theory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vctrs theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

vctrs is heavily inspired by abstract and algebraic data types. The primary goal is not to provide new vector classes (although vctrs does provide a couple), but to think carefully about how existing classes should behave when used together.

This vignette collects together various thoughts related to theory; it's not yet readable.

## Motivating examples

There are two motivating examples from that illustrate many of the key ideas: `ifelse()` and `rbind()`. These have analogs in dplyr (`dplyr::if_else()` and `dplyr::bind_rows()`) which are slightly stricter on average than their base equivalents, but still adhoc. 

What should the result of `ifelse()` look like? `ifelse()` has three inputs: 

* a logical `condition`
* a value to use when the condition is `true`
* a value to use when the condition is `false`

It seems like these three inputs should all effect the output, but precisely how?

`rbind()` poses two challenges. First, what should the overall output type be? i.e. if you combine a data frame and a tibble, what should the result be? What if you combine a data table and a tibble? Second, how should the individual columns be handled, and what should happen when the columns don't match exactly between the inputs?

`rbind()` also poses a challenge of recursion, because the columns of a data frame can themselves [be data frames][df-col]. This means that the type of a data frame is not flat 

## Algebra

`vec_type2()` is a binary operator defined on the set of types. We will call the result of `vec_type2(x, y)` the __commmon type__ of `x` and `y`. The common type is only defined for a small set of pairs. If the common type does exist, we can say that `x` and `y` are uncoercable.

It has three useful properties:

* It is commutative: `vec_type2(x, vec_type(y, z)) == vec_type(vec_type2(x, y), z)`
* It is associative: `vec_type2(x, y) == vec_type2(y, x)`
* It has an identity element, `NULL`: `vec_type2(x, NULL) == x`

Together, this means the type system forms a __commutative monoid__, which has a couple of appealing properties:

* When have multiple types we can find their common type by progressively 
  reducing (aka folding) `vec_type2()`. We'll get the same results regardless
  of whether we reduce from left or right.
  
* The operation of finding common type is [parallelisable][prefix-sum].
  For practial input size, parallelism is unlikely to helpful, but it's 
  nice property to have in general.

## Data frames

Data frames are a __product type__: the type of the data frame is provided by the type of each column tagged with a name.

```{r}
df1 <- data.frame(x = FALSE, y = 2)
vec_type(df1)

df2 <- data.frame(x = 1, y = 2L)
vec_type(df2)
```

The common type of two data frames is the common type of each pair of shared columns:

```{r}
vec_type(vec_type2(df1, df2))
```

By default, the common type of a data frame includes the union of columns in each data frame:

```{r}
df3 <- data.frame(x = 1, y = 2)
df4 <- data.frame(x = 1, z = 3)

vec_type(vec_type2(df3, df4))
```

[df-col]: https://adv-r.hadley.nz/vectors-chap.html#matrix-and-data-frame-columns
[prefix-sum]: https://en.wikipedia.org/wiki/Prefix_sum#Parallel_algorithm
