---
title: "vctrs vs base"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vctrs vs base}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette compares and constrasts existing base R functions with their vctrs equivalents.

```{r}
library(vctrs)
```

## `c()` and `vec_c()`

### Atomic vectors

```{r, error = TRUE}
# c() will coerce any atomic type to character
c(1, "x")

# vctrs is stricter, and requires an explicit cast
vec_c(1, "x")

vec_c(1, "x", .ptype = character())
```

### Factors

```{r}
fa <- factor("a")
fb <- factor("b")

# c() strips all factor attributes giving an integer vector
# (as documented in ?c)
c(fa, fb)

# unlist() creates a new factor with the union of the levels 
fs <- list(fa, fb)
unlist(fs)
```

vctrs always unions the levels. This is a somewhat contentious decision, primarily motivated by pragmatics.

```{r}
vec_c(fa, fb)
```

vctrs doesn't currently provide an `unlist()` equivalent, but you can use a tidy dots feature:

```{r}
vec_c(!!!fs)
```

### Time zones

As documented, `c()` strips time zones:

```{r}
datetime_nz <- as.POSIXct("2020-01-01 09:00", tz = "Pacific/Auckland")

c(datetime_nz)
```

`vec_c()` preserves them:

```{r}
vec_c(datetime_nz)
```

If there are multiple time zones, vctrs takes the first non-local timezone:

```{r}
datetime_local <- as.POSIXct("2020-01-01 09:00")
datetime_houston <- as.POSIXct("2020-01-01 09:00", tz = "US/Central")

vec_ptype(datetime_local, datetime_houston)
vec_ptype(datetime_houston, datetime_nz)
vec_ptype(datetime_nz, datetime_houston)
```

### Dates and date-times

Combining dates and date-times with `c()` gives silently incorrect results:

```{r}
date <- as.Date("2020-01-01")
datetime <- as.POSIXct("2020-01-01 09:00")

# If the first argument to c() is a date, the result is a date
# But the datetime is not converted correctly (the number of seconds
# in the datetime is interpreted as the number of days in the date)
c(date, datetime)

# If the first argument to c() is a datetime, the result is a datetime
# But the date is not converted correctly (the number of days in the
# date is interpreted as the number of seconds in the date)
c(datetime, date)
```

When you mix dates and date-times, vctrs returns a date-time, and converts dates to date-times at midnight (in the timezone of the date-time).

```{r}
vec_c(datetime, date)

vec_c(date, datetime)
vec_c(date, datetime_nz)
```

### Missing values

vctrs treats a logical vector consisting only of `NA` special: it's an unspecified type than can be converted to any other 1d type:

```{r}
vec_c(NA, "x")
vec_c(NA, fa)
vec_c(NA, date)
vec_c(NA, datetime)
```

Whereas `c()` strips the classes:

```{r}
c(NA, "x")
c(NA, fa)
c(NA, date)
c(NA, datetime)
```

## `rbind()` and `vec_rbind()`

### Variables

`rbind()` requires all inputs to have the same variables

```{r, error = TRUE}
df1 <- data.frame(x = TRUE)
df2 <- data.frame(y = 2)

# rbind() requires the inputs to have identical column names
rbind(df1, df2)
```

`vec_rbind()` takes the union of columns:

```{r}
vec_rbind(df1, df2)
```

### Mixing data frames and vectors

`rbind()` can silently discard data:

```{r}
rbind(data.frame(x = 1:3), c(1, 1000000))
```

`vec_rbind()` first converts any vector inputs to one-row data frames.

```{r}
vec_rbind(data.frame(x = 1:3), c(1, 1000000))
```

### Container class

`rbind()` uses the class of the first argument:

```{r}
tb1 <- tibble::tibble(x = 3)

class(rbind(tb1, df1))
class(rbind(df1, tb1))
```

`vec_rbind()` uses the common class:

```{r}
class(vec_rbind(df1, df1))
class(vec_rbind(tb1, df1))
class(vec_rbind(df1, tb1))
```

Like `vec_c()` you can specify the desired output type:

```{r}
vec_rbind(df1, df2, .ptype = data.frame(x = double(), y = double()))
```

### Empty inputs

Because `rbind()` can produce both matrices and data frames, its identity element is `NULL`:

```{r}
rbind()
rbind(NULL)
```

Because `vec_rbind()` only produces data frames, its identity element is an empty data frame:

```{r}
vec_rbind()
vec_rbind(NULL)
```

