
```{r, child = "../setup.Rmd", include = FALSE}
```

This guide provides a practical recipe for implementing `vec_ptype2()` and `vec_cast()` methods for coercions of data frame subclasses. Related topics:

- For an overview of the coercion mechanism in vctrs, see [`?theory-faq-coercion`][theory-faq-coercion].

- For an example of implementing coercion methods for non-data-frame vectors, see [`?howto-faq-coercion`][howto-faq-coercion].

Coercion of data frames occurs when different data frame classes are combined in some way. The two main methods of combination are currently row-binding with [vec_rbind()] and col-binding with [vec_cbind()] (which are in turn used by a number of dplyr functions). These functions take multiple data frame inputs and automatically coerce them to their common type.

vctrs is generally strict about the kind of automatic coercions that are performed when combining inputs. In the case of data frames we have decided to be a bit less strict for convenience. Instead of throwing an incompatible type error, we fall back to  a base data frame or a tibble if we don't know how to combine two data frame subclasses. It is still a good idea to specify the proper coercion behaviour for your data frame subclasses as soon as possible.


## Parent methods

Most of the common type determination should be performed by the parent class. In vctrs, double dispatch is implemented in such a way that you need to call the methods for the parent class manually. For `vec_ptype2()` this means you need to call `df_ptype2()` (for data frame subclass) or `tib_ptype2()` (for tibble subclass). Similarly, `df_cast()` and `tib_cast()` are the workhorses for `vec_cast()` methods of subtypes of `data.frame` and `tbl_df`. These functions take the union of the columns in `x` and `y`, and ensure shared columns have the same type.

These functions are much less strict than `vec_ptype2()` and `vec_cast()` as they accept any subclass of data frame as input. They always return a `data.frame` or a `tbl_df`. Note that if you write similar functions for your subclass (which should be rare), you'll need to export it so.


## A `data.table` example

```{r, include = FALSE}
delayedAssign("as.data.table", {
  if (is_installed("data.table")) {
    env_get(ns_env("data.table"), "as.data.table")
  } else {
    function(...) abort("`data.table` must be installed.")
  }
})
delayedAssign("data.table", {
  if (is_installed("data.table")) {
    env_get(ns_env("data.table"), "data.table")
  } else {
    function(...) abort("`data.table` must be installed.")
  }
})
```

This example is the actual implementation of vctrs coercion methods for `data.table`.

We first create the `dt_ptype2()` and `dt_cast()` helpers. They wrap around the parent methods `df_ptype2()` and `df_cast()`, and transform the common type or converted input to a data table. You may want to export these helpers if you expect other packages to derive from your data frame class.

These helpers should always return data tables. To this end we use the conversion generic `as.data.table()`. Depending on the tools available for the particular class at hand, a constructor might be appropriate as well.

```{r}
dt_ptype2 <- function(x, y, ...) {
  as.data.table(df_ptype2(x, y, ...))
}
dt_cast <- function(x, to, ...) {
  as.data.table(df_cast(x, to, ...))
}
```

We start with the self-self method:

```{r}
#' @export
vec_ptype2.data.table.data.table <- function(x, y, ...) {
  dt_ptype2(x, y, ...)
}
```

Between a data frame and a data table, we consider the richer type to be data table. This is not made based on the coverage of each data structures, but on the idea that data tables have richer behaviour. Since data tables are the richer type, we call `dt_type2()` from the `vec_ptype2()` method. It always returns a data table, no matter the order of arguments:

```{r}
#' @export
vec_ptype2.data.table.data.frame <- function(x, y, ...) {
  dt_ptype2(x, y, ...)
}
#' @export
vec_ptype2.data.frame.data.table <- function(x, y, ...) {
  dt_ptype2(x, y, ...)
}
```

The `vec_cast()` methods follow the same pattern, but note how the method for coercing to data frame uses `df_cast()` rather than `dt_cast()`:

```{r}
#' @export
vec_cast.data.table.data.table <- function(x, to, ...) {
  dt_cast(x, to, ...)
}
#' @export
vec_cast.data.table.data.frame <- function(x, to, ...) {
  dt_cast(x, to, ...)
}
#' @export
vec_cast.data.frame.data.table <- function(x, to, ...) {
  df_cast(x, to, ...)
}
```

With these methods vctrs is now able to combine data tables with data frames:

```{r}
vec_cbind(data.frame(x = 1:3), data.table(y = "foo"))
```


## A tibble example

In this example we implement coercion methods for a tibble subclass:

```{r}
# User constructor
my_tibble <- function(...) {
  new_my_tibble(tibble::tibble(...))
}
# Developer constructor
new_my_tibble <- function(x) {
  stopifnot(is.data.frame(x))
  tibble::new_tibble(x, class = "my_tibble", nrow = nrow(x))
}

print.my_tibble <- function(x, ...) {
  cat("<my_very_own_tibble_class>\n")
  cli::cat_line(format(x)[-1])
}
```

This subclass is very simple. All it does is modify the header.

```{r}
df <- my_tibble(x = 1, y = 1:2)

df

df[2]
```

Combinations do not work properly out of the box, instead vctrs falls back to a bare tibble:

```{r}
vec_rbind(df, tibble(x = 10:12))
```

We would like to return a `<my_tibble>` when combined with a data frame or a tibble. This subclass of tibble is in fact a supertype of tibble, i.e. it is the richer type.

We'll need to call our parent methods [tib_ptype2()] and [tib_cast()] and pass their results to our constructor. This is best wrapped in two functions to avoid repetition:

```{r}
#' @export
my_tib_cast <- function(x, to, ...) {
  new_my_tibble(tib_cast(x, to, ...))
}
#' @export
my_tib_ptype2 <- function(x, y, ...) {
  new_my_tibble(tib_ptype2(x, y, ...))
}
```

Let's implement the coercion methods, starting with the self-self methods.

```{r}
#' @export
vec_ptype2.my_tibble.my_tibble <- function(x, y, ...) {
  my_tib_ptype2(x, y, ...)
}
#' @export
vec_cast.my_tibble.my_tibble <- function(x, y, ...) {
  my_tib_cast(x, y, ...)
}
```

The methods for combining our class with tibbles follow the same pattern. For ptype2 we return our class in both cases because it is the richer type:

```{r}
#' @export
vec_ptype2.my_tibble.tbl_df <- function(x, y, ...) {
  my_tib_ptype2(x, y, ...)
}
#' @export
vec_ptype2.tbl_df.my_tibble <- function(x, y, ...) {
  my_tib_ptype2(x, y, ...)
}
```

For cast are careful about returning a tibble when casting to a tibble. Note the call to `vctrs::tib_cast()`:

```{r}
#' @export
vec_cast.my_tibble.tbl_df <- function(x, y, ...) {
  my_tib_cast(x, y, ...)
}
#' @export
vec_cast.tbl_df.my_tibble <- function(x, y, ...) {
  tib_cast(x, y, ...)
}
```

From this point, we get correct combinations with tibbles:

```{r}
vec_rbind(df, tibble(x = 10:12))
```

However we are not done yet. Because the coercion hierarchy is different from the class hierarchy, there is no inheritance of coercion methods. We're not getting correct behaviour for data frames yet:

```{r}
vec_rbind(df, data.frame(x = 10:12))
```

Let's finish up the boiler plate:

```{r}
#' @export
vec_ptype2.my_tibble.data.frame <- function(x, y, ...) {
  tib_ptype2(x, y, ...)
}
#' @export
vec_ptype2.data.frame.my_tibble <- function(x, y, ...) {
  tib_ptype2(x, y, ...)
}

#' @export
vec_cast.my_tibble.data.frame <- function(x, y, ...) {
  tib_cast(x, y, ...)
}
#' @export
vec_cast.data.frame.my_tibble <- function(x, y, ...) {
  df_cast(x, y, ...)
}
```

This completes the implementation:

```{r}
vec_rbind(df, data.frame(x = 10:12))
```
